---
title: "Appendix 1 - The Swedish Covid-19 Strategy: A Comparison of the National Policy Response between the First and Second Waves"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```

This notebook reports the sources and transformations used to generate the descriptive data reported in the study "The Swedish Covid-19 Strategy: A Comparison of the National Policy Response between the First and Second Waves". The source code and raw data are available at https://github.com/dnspangler/hepl_covid_2020

# Load and transform data

```{r load}

library(readxl)
library(tidyverse)

options(scipen = 999)

# Load national Covid-19 tracking data from:
# https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/statistik-och-analyser/bekraftade-fall-i-sverige/

# Specifically:
# https://www.arcgis.com/sharing/rest/content/items/b5e7488e117749c19881cce45db13f7e/data

fhm <- read_excel("Folkhalsomyndigheten_Covid19.xlsx", 
                                           sheet = "Veckodata Region") %>%
  mutate(date = as.Date(paste(2020, veckonummer, 1, sep="-"), "%Y-%U-%u"))


# Load testing data from PHA - This is for some reason not included in the main dataset, and had to be manually compiled from:
# https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/statistik-och-analyser/antalet-testade-for-covid-19/
# https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/statistik-och-analyser/antalet-testade-for-covid-19/tidigare-data/

fhm_test <- read_excel("fhm_Covid19_tester.xlsx")


# Load data from Statistics Sweden on mortality rates from 2015 - 2020
#https://www.scb.se/hitta-statistik/statistik-efter-amne/befolkning/befolkningens-sammansattning/befolkningsstatistik/pong/tabell-och-diagram/preliminar-statistik-over-doda/

scb <- read_excel("preliminar_statistik_over_doda_inkl_eng.xlsx", 
                  sheet = "Tabell 6",skip = 11) %>%
  mutate(veckonummer = as.numeric(Vecka))

# Load data from NBHW from nbhw - Unfortunately looks like this isn't getting updated anymore!
#https://www.socialstyrelsen.se/globalassets/1-globalt/covid-19-statistik/statistik-relaterad-till-covid-19/dodlighet-aldre-med-sarskilt-boende-eller-hemtjanst.xlsx

nbhw <- read_excel("dodlighet-aldre-med-sarskilt-boende-eller-hemtjanst.xlsx", 
                  sheet = "Jämf genomsnitt, antal",
                  skip = 3) %>%
  filter(!is.na(`Särskilt boende (2020)`)) %>%
  mutate(excess_nh = `Särskilt boende (2020)`- `Särskilt boende (genomsnitt 2016-2019)`,
         excess_hc = `Hemtjänst (2020)`- `Hemtjänst (genomsnitt 2016-2019)`,
         excess_no_nh_hc = `Ej särskilt boende eller hemtjänst (2020)` - `Ej särskilt boende eller hemtjänst (genomsnitt 2016-2019)`,
         excess_over_70 = excess_nh+excess_hc+excess_no_nh_hc,
         tot_over70 = `Särskilt boende (2020)`+`Hemtjänst (2020)`+`Ej särskilt boende eller hemtjänst (2020)`,
         veckonummer = as.numeric(Vecka),
         date = as.Date(paste(2020, veckonummer, 1, sep="-"), "%Y-%U-%u"))

# Combine scb and nbhw data to calculate mortality under 70
excess_mort = select(scb,veckonummer,excess_all = `Överdödl. Riket`,tot = `Tot 2020`) %>%
  left_join(select(nbhw,veckonummer,date,starts_with("excess_"),tot_over70), by = "veckonummer") %>%
  mutate(excess_under_70 = excess_all - excess_over_70) %>%
  filter(veckonummer > 1) # Looks like scb only counts part of the first week of the year, So not comparable

# aggregate data at national level

agg <- fhm %>%
  group_by(veckonummer) %>%
  summarize(across(where(is.numeric),sum)) %>%
  ungroup() %>%
  left_join(fhm_test, by = "veckonummer") %>%
  left_join(select(scb,veckonummer, `Medelvärde 2015-2019`,`Tot 2020`,`Överdödl. Riket`),
            by = "veckonummer") %>%
  mutate(date = as.Date(paste(2020, veckonummer, 1, sep="-"), "%Y-%U-%u"),
         test_pos_rate = Antal_fall_vecka/Antal_pcr*100,
         case_fatality_rate = Antal_avlidna_vecka/Antal_fall_vecka*100)

# Dataset grouped by region

fhm_reg_sum <- fhm %>%
  group_by(Region) %>%
  summarise(tot = sum(Antal_fall_vecka),
            pop = round(sum(Antal_fall_vecka)/sum(Antal_fall_100000inv_vecka)*100000)) %>%
  arrange(desc(tot)) %>%
  mutate(agg_name = ifelse(tot > 3000,Region,"Other"),
         agg_name = factor(agg_name,levels = rev(unique(agg_name))))

```

## Figure 1

```{r fig1,fig.width=4,fig.height=8}
# FHM stats aggregated
agg_plot <- agg %>%
  select(date,
         "Covid-19 PCR tests performed" = Antal_pcr,
         "New Covid-19 cases" = Antal_fall_vecka,
         "New Covid-19 cases in ICU" = Antal_intensivvårdade_vecka,
         "Covid-19-related Mortalities" = Antal_avlidna_vecka,
         "Excess mortality (2020 v 2015-2019)" = `Överdödl. Riket`,
         "Test positivity rate (%)" = test_pos_rate,
         "Case fatality rate (%)" = case_fatality_rate) %>%
  gather("measure","value",-date) %>%
  # remove last week of mortality data - The second to last week of data also 
  # tends to under-report by a smaller amount, but not enough to significantly 
  # skew the interpretation.
  filter(!(measure %in% c("Covid-19-related Mortalities",
                          "Excess mortality (2020 v 2015-2019)",
                          "Case fatality rate (%)") & 
             date==max(date))) %>% 
  mutate(measure = factor(measure,levels = unique(measure))) %>%
  ggplot(aes(x=date,y = value)) +
  theme_minimal() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  theme(panel.grid.minor.x = element_blank()) +
  geom_hline(yintercept = 0,color = "grey") +
  geom_line(size = 1) +
  #scale_y_continuous(breaks = seq(-200,800,200)) +
  facet_wrap(~ measure,scales = "free_y",ncol = 1)


# Use this to export a nice high-res .tif file
# tiff("agg_plot.tif", res=300,height = 8,width = 4,units = "in")
# agg_plot
# dev.off()

agg_plot

```

## Additional figures

Due to space constraints, we had to cut some interesting stuff. Here are a couple of plots which didn't make it into the final paper:

### Per capita cases and fatalities per region
```{r reg_plot,fig.width=4,fig.height=8}

reg_plot <- fhm %>% 
  left_join(fhm_reg_sum, by = "Region") %>%
  arrange(desc(pop)) %>%
  mutate(Region = factor(Region,levels = unique(Region)),
         Antal_avliden_1000000inv_vecka = Antal_avlidna_vecka/(pop/1000000)) %>%
  select(date,
         Region,
         "Cases per 100k" = Antal_fall_100000inv_vecka,
         "Deaths per million" = Antal_avliden_1000000inv_vecka) %>%
  gather("measure","value",-date,-Region) %>%
  mutate(measure = factor(measure,levels = rev(unique(measure)))) %>%
  ggplot(aes(x=date,y=value,color=measure)) + 
  geom_smooth(se=F,span = 0.2,size = 1) +
  geom_point(size=0.1) +
  theme_minimal() +
  labs(color = "Covid related...") +
  scale_y_continuous(breaks = c(0,100,200)) +
  scale_x_date(date_breaks = "2 months",
               date_minor_breaks = "1 month", 
               date_labels = "%b") +
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = c(0.75, 0.01)) +
  facet_wrap(Region ~ .,ncol = 2,)

reg_plot
```

### Excess mortality by age and form of social services

```{r}

excess_mort %>%
  filter(!is.na(excess_nh)) %>%
  arrange(desc(veckonummer)) %>%
  mutate(date = as.Date(paste(2020, veckonummer, 1, sep="-"), "%Y-%U-%u")) %>%
  select(date,
         "70+ in nursing home" = excess_nh,
         "70+ with home care" = excess_hc,
         "70+ with no social services" =excess_no_nh_hc,
         "<70 years old"=excess_under_70) %>%
  gather("measure","value",-date) %>%
  mutate(measure = factor(measure,levels = unique(measure))) %>%
  ggplot(aes(x=date,y=value,fill=measure)) +
    scale_y_continuous(breaks = seq(-200,800,100)) +
    scale_x_date(date_breaks = "2 months",
                 date_minor_breaks = "1 month", 
                 date_labels = "%b") +
    scale_fill_brewer(palette = "Set2") +
    #scale_color_brewer(palette = "Set2") +
    labs(y = "Excess mortality per week vs. 2015-2019",
         fill = "Population cohort") +
    geom_bar(stat = "identity",width = 7)
```

